A Quantitative Genetic Linked Selection Model of the Negative Selection Process in Humans

### 2023-07-23 manuscript/manuscript.tex
ms updates from italy

    Across the human genome, there are large-scale fluctuations in genetic
    diversity caused by the indirect effects of selection. This can be thought
    of as a ``linked selection signal" that reflects the effects of selection
    varying according to the placement of functional regions and recombination
    rates along the genome. Previous work has shown that negative selection
    against the steady influx of new deleterious mutations into conserved
    regions is the predominant mode of selection in humans. However, the
    theoretic model that underpins these results, classic Background Selection
    theory, is only applicable when new mutations are so deleterious that they
    cannot drift up in frequency and ultimately fix in the population. Here, we
    develop and apply a statistical method to fit the linked selection signal
    based on an alternate theoretic approach. This theory is based on a
    quantitative genetics view of the linked selection process, which models
    the cumulative effects of weak draft created according the arrangement of
    polygenic additive fitness variance along the genome. Using a recent model
    that jointly predicts the equilibrium fitness variance and substitution
    rates due to deleterious mutations, we estimate the distribution of fitness
    effects (DFE) and mutation rate across three human populations. While our
    model can accommodate weaker selection, we initially find evidence across
    three human populations of very strong selection against deleterious
    mutations consistent with previous work. However, the corollary predictions
    of the substitution rates from our model suggest an unreasonably low rate
    of divergence in highly-conserved region which disagree with observations.
    We hypothesize that this is driven by selected sites within conserved
    segments experiencing a reduced effective population size due to selection
    across other segments. Extending our method to approximate these effects
    through a locally-reduced effective population size brings predicted
    substitution rates into agreement with observed divergence rates and
    revises DFE estimates to include a sizable portion of weakly deleterious
    mutations. However, this also leads to implausibly large mutation rate
    estimates. Overall, while our model of the genomic linked selection signal
    brings us a step towards uniting population and quantitative genetic
    selection models with the substitution process, our work suggests
    considerable uncertainty remains about the processes generating fitness
    variance in humans.
\end{abstract}

@@

Prior work to model the reduction in linked diversity due to deleterious
mutations has largely relied on the classic Background Selection (BGS) model
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1995-xc}.
While this model has been successful in explaining many patterns of diversity,
some its simplifying assumptions may distort inferences about the negative
selection process. First, since fixation probabilities ultimately depend on the
product of the deleterious selection coefficient ($s$) and population size
($N$), the efficacy of negative selection should depend on demography.
Unfortunately, accommodating demography into theoretic negative selection
models remains an open, difficult problem \parencite{Zeng2013-ep,Johri2020-oj}.
Second, building off classic models of mutation-selection balance
\parencite{Crow1970-wm,Kimura1966-bk}, the BGS model assumes that new mutations
are sufficiently deleterious that they are invariably driven to loss. Under
this assumption, the effect of selection is well-approximated by simply
rescaling the neutral coalescent by a reduction factor known as $B =
\nicefrac{N}{N_e}$ \parencite{Charlesworth2013-kl}. However, this simple
rescaling approach is not appropriate across parts of parameter space that are
relevant to natural populations \parencite{McVean2000-bt,Good2014-yz}. In
particular, the BGS model cannot accommodate the possibility of fixation for
mutations with $2Ns \ll 1$, leading diversity to be incorrectly predicted to
fall to zero as the strength of selection diminishes. This weak selection
problem is related to the wider problem of modeling the substitution rate of
deleterious mutation (i.e. the ``ratchet" rate) and incorporating the impact of
interference from other neighboring selected sites. Finally, the classic BGS
model assumes that selection occurring within other segments does not interfere
with selection occurring in other regions, i.e. no selective or
``Hill--Robertson" interference
\parencite{Hill1966-kd,McVean2000-bt,Felsenstein1974-xm}.

In this work, we use another class of linked selection models that derive from
quantitative genetics to address limitations of the classic BGS model. These
models express linked selection in terms of how polygenic fitness variance
creates weak genetic draft, and which increases pairwise coalescence rates.
While these models can theoretically accommodate additive fitness variance from
any source as long as its rate of change is not too rapid, we focus
specifically on a deleterious-mutations-only model of fitness variance from
\textcite{Santiago2016-mu}. This model is identical to BGS when selection
against deleterious mutations is strong, but it also correctly predicts the
reduction in diversity when selection is weak by jointly predicting the
deleterious substitution rate. We extend the Santiago and Caballero (hereafter
the SC16) model of the negative selection process so that it can be fit to
patterns of genome-wide diversity, according to the spatial distribution of
genomic features that could harbor deleterious fitness variation. Using forward
simulations, we show this model leads to more accurate estimates of the
distribution of fitness effects under weak selection. Finally, we apply our
composite-likelihood method to human population genomic data and provide new
estimates of the negative selection process in humans.

@@

The remaining required expressions are the equilibrium additive fitness
variance $V_A$ and the decay rate in associated fitness $1-\kappa$. Fitness
variance could arise from positively or negatively selected alleles, but given
prior work has found negative selection against new deleterious mutations play
a dominant role in shaping genome-wide patterns of diversity
\parencite{McVicker2009-ax,Murphy2022-sj}, we focus on models of fitness
variance created by the continual influx of deleterious mutations into a
population. We imagine a mutation-selection process that creates fitness
variance through a continual influx of deleterious mutations occur at a rate of
$\mu$ per basepair per generation in a conserved region of $L$ basepairs, such
that the region-wide per generation diploid mutation rate is $U = 2 \mu L$.
Each mutation imposes a selective cost of $s$ in heterozygotes and $2s$ in
homozygotes, and fitness effects are multiplicative across sites.

@@

However, when new mutations are only weakly deleterious, they can drift up in
frequency before their eventual loss or fixation. At this point, the number of
deleterious mutations is no longer well-approximated by the deterministic
mutation-selection-balance theory; their dynamics are strongly influenced by
stochastic perturbations due to both drift and draft. In this weak selection
regime, classic BGS no longer accurately predicts levels of linked diversity
\parencite{Charlesworth1993-gb,McVean2000-bt,Good2013-lp,Gordo2002-dr}.
Moreover, selection against weakly deleterious mutations alters the
genealogies, such that they are no longer well-approximated by a rescaled
neutral coalescent as under the classic BGS model
\parencite{Przeworski1999-mb,OFallon2010-my,Higgs1995-xc}. To further
complicate matters, the distribution of the number of deleterious mutations
(and its corresponding fitness distribution) is no longer a stationary Poisson
distribution, instead becoming a traveling wave
\parencite{Rouzine2008-qz,Good2013-lp,Gessler1995-hz} towards reduced
population fitness (since we ignore beneficial mutations). When this occurs,
individuals in the least-loaded class are stochastically lost, fixing those
mutations with a click of ``Muller's ratchet"
\parencite{Muller1964-ki,Charlesworth1997-qn}. Unfortunately, determining the
rate of Muller's ratchet is another difficult, open problem
\parencite{Haigh1978-gt,Gordo2002-dr,Gessler1995-hz,Higgs1995-xc} related to
Hill--Robertson interference \parencite{Felsenstein1974-xm}.

Under qualitative genetic models of linked selection, these complications can
can be avoided by finding an expression for the equilibrium fitness variance
that is correct for both strong and weak deleterious mutations.

\textcite{Santiago2016-mu} suggested that this is due to the loss of fitness
variation that occurs when a segregating site fixes and its heterozygosity
becomes zero. If we let $R$ represent the fixation rate (i.e. ratchet rate) in
the region per generation, each fixation removes the equivalent amount of
equilibrium variation put in by mutation. Thus, the steady-state genic variance
under mutation and negative selection is (omitting the segment index $i$ for
clarity),

@@

Given the accuracy of the SC16 model in predicting the reduction factor $B$ and
ratchet rate for a single segment under general mutation-selection processes,
we next extended their model so that it could be applied to whole-genome data.
Our software method \texttt{bgspy} numerically solves Equations
\eqref{eq:main_eqns} to compute the equilibrium additive genic fitness variance
($V_a$) and ratchet rate ($R$) across grids of mutation rates and selection
coefficients. This is done for each pre-specified segment in the genome that
may be under negative selection (e.g. coding sequences or UTRs). Then, these
equilibrium can be used to calculate a reduction map $B(x)$ across genomic
positions $x$. To distinguish between McVicker's B maps based on classic BGS
theory, we call our reduction map $B(x)$ a B' map.

@@

The discrepancy between our theoretic B' maps and chromosome-wide simulations
that leads to the 5\% error occurs entirely around the drift-barrier domain of
$2Ns \approx 1$ (Figure \ref{fig:figure-1}D). The higher error in this domain
occurs across all mutation rates. We hypothesized this error may be due to
selective interference between segments that is not taken into account when we
numerically solve Equations \eqref{eq:main_eqns} for each segment
independently. In particular, we use a fixed drift-effective $N=1000$
corresponding to the number of diploids in the simulations rather than $B(x) N$
at position $x$. To test this, we implemented a ``locally-rescaled" version of
the B' maps, which numerically solves Equations \eqref{eq:main_eqns} for each
segment at position $x$ taking into account the reduction experienced due to
selection at other segments by setting $B'(x)N$ as the population size. We find
the locally-rescaled B' maps reduce the relative error from 5\% to 0.4\% and
mean squared error (Figure \ref{fig:figure-1}D, dashed colored lines), but does
not entirely eliminate the error in the $2Ns \approx 1$ domain. Overall, while
we find evidence that local-rescaling slightly reduces error in the $2Ns
\approx 1$ domain, we do not use it during in our genome-wide inference method.
This is because to avoid circularity and possible overfitting, our method would
need to solve the equilibria equations for all segments simultaneously;
unfortunately, this is currently computational unfeasible.

@@

We conducted forward simulations of negative selection on whole human
chromosomes to validate our method at two stages. First, we simulated negative
selection on chromosome 10 using a realistic recombination map and putatively
conserved features to confirm that our classic B and new B' maps matched the
average simulation reduction map across mutation and selection parameters.
Second, we evaluated our composite likelihood method by simulating negative
selection on the first five human chromosomes, across grids of fixed mutation
and selection parameters. We then combined these into a synthetic genome, and
overlaid mutations on the ARG. Then, we ran our likelihood methods on the
resulting allele count data to assess model accuracy. We ran additional
synthetic genome simulations like these to evaluate the impact of two model
violations: recessivity of deleterious mutations and expanding populations. For
the latter, after $9.3$ generations, we grew the population by factor of 1.004
each generation to mimic the human expansion out-of-Africa
\parencite{Gutenkunst2009-pg}. We did not simulate population bottlenecks since
our analyses showed little difference between bottlenecked out-of-Africa
samples (CEU and CHB) and YRI samples. More details about these and the segment
simulations shown in Figure \ref{fig:figure-1}A-C can be found Supplementary
Section \ref{supp:simulation}.

>>>>

    Across the human genome, there are large-scale fluctuations in genetic
    diversity caused by the indirect effects of selection. This can be thought
    of as a ``linked selection signal" that reflects the effects of selection
    varying according to the placement of functional regions and recombination
    rates along the genome. Previous work has shown that negative selection
    against the steady influx of new deleterious mutations into conserved
    regions is the predominant mode of selection in humans. However, the
    theoretic model that underpins these results, classic Background Selection
    theory, is only applicable when new mutations are so deleterious that they
    cannot drift up in frequency and ultimately fix in the population. Here, we
    develop and apply a statistical method to fit the linked selection signal
    based on an alternate theoretic approach. This theory is based on a
    quantitative genetics view of the linked selection process, which models
    the cumulative effects of weak draft created according the arrangement of
    polygenic additive fitness variance along the genome. Using a recent model
    that jointly predicts the equilibrium fitness variance and substitution
    rates due to deleterious mutations, we estimate the distribution of fitness
    effects (DFE) and mutation rate across three human populations. While our
    model can accommodate weaker selection, we initially find evidence across
    three human populations of very strong selection against deleterious
    mutations consistent with previous work. However, the corollary predictions
    of the substitution rates from our model suggest an unreasonably low rate
    of divergence in highly-conserved region which disagree with observations.
    We hypothesize that this is driven by selected sites within conserved
    segments experiencing a further reduced effective population size due to
    selection across other segments. Extending our method to approximate these
    effects through a locally-reduced effective population size brings
    predicted substitution rates into agreement with observed divergence rates
    and revises DFE estimates to include a sizable portion of weakly
    deleterious mutations. However, this also leads to implausibly large
    mutation rate estimates. Overall, while our model of the genomic linked
    selection signal brings us a step towards uniting population and
    quantitative genetic selection models with the substitution process, our
    work suggests considerable uncertainty remains about the processes
generating fitness variance in humans.

\end{abstract}

@@

Prior work to model the reduction in linked diversity due to deleterious
mutations has largely relied on the classic Background Selection (BGS) model
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1995-xc}.
While this model has been successful in explaining many patterns of diversity,
some its simplifying assumptions may distort inferences about the negative
selection process. First, since fixation probabilities ultimately depend on the
product of the deleterious selection coefficient ($s$) and population size
($N$), the efficacy of negative selection should depend on demography.
Unfortunately, accommodating demography into theoretic negative selection
models remains an open, difficult problem \parencite{Zeng2013-ep,Johri2020-oj}.
Second, building off classic models of mutation-selection balance
\parencite{Crow1970-wm,Kimura1966-bk}, the BGS model assumes that new mutations
are sufficiently deleterious that they are invariably driven to loss. Under
this assumption, the effect of selection is well-approximated by simply
rescaling the neutral coalescent by a reduction factor known as $B =
\nicefrac{N_e}{N}$ \parencite{Charlesworth2013-kl}. However, this simple
rescaling approach is not appropriate across parts of parameter space that are
relevant to natural populations \parencite{McVean2000-bt,Good2014-yz}. In
particular, the BGS model cannot accommodate the possibility of fixation for
mutations with $2Ns \ll 1$, leading diversity to be incorrectly predicted to
fall to zero as the strength of selection diminishes. This weak selection
problem is related to the wider problem of modeling the substitution rate of
deleterious mutation (i.e. the ``ratchet" rate) and incorporating the impact of
interference from other neighboring selected sites. Finally, the classic BGS
model assumes that selection occurring within other segments does not interfere
with selection occurring in other regions, i.e. no selective or
``Hill--Robertson" interference
\parencite{Hill1966-kd,McVean2000-bt,Felsenstein1974-xm}.

In this work, we use another class of linked selection models that derive from
quantitative genetics to address limitations of the classic BGS model. These
models express linked selection in terms of how polygenic fitness variance
creates weak genetic draft, and which increases pairwise coalescence rates.
While these models can theoretically accommodate additive fitness variance from
any source as long as its rate of change is not too rapid, we focus
specifically on a deleterious-mutations-only model of fitness variance from
\textcite{Santiago2016-mu}. This model is identical to BGS when selection
against deleterious mutations is strong, but it also correctly predicts the
reduction in diversity when selection is weak by jointly predicting the
deleterious substitution rate. We extend the Santiago and Caballero (hereafter
the SC16) model of the negative selection process so that it can be fit to
patterns of genome-wide diversity, according to the spatial distribution of
genomic features that could harbor deleterious fitness variation. Using forward
simulations, we show this model leads to more accurate estimates of the
distribution of fitness effects under weak selection. We apply our
composite-likelihood method to human population genomic data and provide new
estimates of the negative selection process in humans. However, we find that
there may be considerable uncertainty around these estimates if selective
interference is occurring across segments under purifying selection.

@@

The remaining required expressions are for the equilibrium additive fitness
variance $V_A$ and the decay rate in associated fitness $1-\kappa$. Fitness
variance could arise from positively or negatively selected alleles, but given
prior work has found negative selection against new deleterious mutations play
a dominant role in shaping genome-wide patterns of diversity
\parencite{McVicker2009-ax,Murphy2022-sj}, we focus on models of fitness
variance created by the continual influx of deleterious mutations into a
population. We imagine a mutation-selection process that creates fitness
variance through a continual influx of deleterious mutations occur at a rate of
$\mu$ per basepair per generation in a conserved region of $L$ basepairs, such
that the region-wide per generation diploid mutation rate is $U = 2 \mu L$.
Each mutation imposes a selective cost of $s$ in heterozygotes and $2s$ in
homozygotes, and fitness effects are multiplicative across sites.

@@

However, when new mutations are only weakly deleterious, they can drift up in
frequency before their eventual loss or fixation. At this point, the number of
deleterious mutations is no longer well-approximated by the deterministic
mutation-selection-balance theory; their dynamics are strongly influenced by
stochastic perturbations due to both drift and draft. In this weak selection
regime, classic BGS no longer accurately predicts levels of linked diversity
\parencite{Charlesworth1993-gb,McVean2000-bt,Good2013-lp,Gordo2002-dr}.
Moreover, selection against weakly deleterious mutations alters the
genealogies, such that they are no longer well-approximated by a rescaled
neutral coalescent as assumed under the classic BGS model
\parencite{Przeworski1999-mb,OFallon2010-my,Higgs1995-xc}. To further
complicate matters, the distribution of the number of deleterious mutations
(and its corresponding fitness distribution) is no longer a stationary Poisson
distribution, instead becoming a traveling wave
\parencite{Rouzine2008-qz,Good2013-lp,Gessler1995-hz} towards reduced
population fitness (since we ignore beneficial mutations). When this occurs,
individuals in the least-loaded class are stochastically lost, fixing those
mutations with a click of ``Muller's ratchet"
\parencite{Muller1964-ki,Charlesworth1997-qn}. Unfortunately, determining the
rate of Muller's ratchet is another difficult, open problem
\parencite{Haigh1978-gt,Gordo2002-dr,Gessler1995-hz,Higgs1995-xc} related to
Hill--Robertson interference \parencite{Felsenstein1974-xm}.

Under qualitative genetic models of linked selection, these complications are
avoided by finding a general expression for the equilibrium fitness variance
that is applicable for both strong and weakly deleterious mutations.
\textcite{Santiago2016-mu} suggest that a general equation for equilibrium
fitness variance must account for the chance fixations of deleterious mutations
that can occur. Each fixation (or click of Muller's ``ratchet") decreases the
fitness variance, as a previously segregating polymorphism fixes and its
heterozygosity becomes zero. If we let $R$ represent the fixation rate (i.e.
ratchet rate) in the region per generation, each fixation removes the
equivalent amount of equilibrium variation put in by mutation. Thus, the
steady-state genic variance under mutation and negative selection is (omitting
the segment index $i$ for clarity),

@@

Given the accuracy of the SC16 model in predicting the reduction factor $B$ and
ratchet rate for a single segment under general mutation-selection processes,
we next extended their model so that it could be applied to whole-genome data.
Our software method \texttt{bgspy} numerically solves Equations
\eqref{eq:main_eqns} to compute the equilibrium additive genic fitness variance
($V_a$) and ratchet rate ($R$) across grids of mutation rates and selection
coefficients. This is done for each pre-specified segment in the genome that
may be under negative selection (e.g. coding sequences or UTRs). We call the
set of predicted reductions across these grids the B' maps (to distinguish them
from McVicker's B maps); these can be used then to find the equilibrium
reduction factor $B(x)$ for any genomic position $x$.

@@

The discrepancy between our theoretic B' maps and chromosome-wide simulations
that leads to the 5\% error occurs entirely around the drift-barrier domain of
$2Ns \approx 1$ (Figure \ref{fig:figure-1}D). The higher error in this domain
occurs across all mutation rates. We hypothesized this error may be due to
selective interference between segments that is not taken into account when we
numerically solve Equations \eqref{eq:main_eqns} for each segment
independently. In particular, we use a fixed drift-effective $N=1000$
corresponding to the number of diploids in the simulations rather than $B(x)
N$, which is an approximation to the draft-effective population size
experienced by a selected segment at position $x$. To test this, we implemented
a ``locally-rescaled" version of the B' maps, which numerically solves
Equations \eqref{eq:main_eqns} for each segment at position $x$ taking into
account the reduction experienced due to selection at other segments by setting
$B(x)N$ as the population size. We find the locally-rescaled B' maps reduce the
relative error from 5\% to 0.4\% and mean squared error (Figure
\ref{fig:figure-1}D, dashed colored lines), but does not entirely eliminate the
error in the $2Ns \approx 1$ domain. Overall, while we find evidence that
local-rescaling slightly reduces error in the $2Ns \approx 1$ domain, we do not
use it during in our genome-wide inference method. This is because to avoid
circularity and possible overfitting, our method would need to solve the
equilibria equations for all segments simultaneously; unfortunately, this is
currently computational unfeasible.

@@

We conducted forward simulations of negative selection on whole human
chromosomes to validate our method at two stages. First, we simulated negative
selection on chromosome 10 using a realistic recombination map and putatively
conserved features to confirm that our classic B and new B' maps matched the
average simulation reduction map across mutation and selection parameters.
Second, we evaluated our composite likelihood method by simulating negative
selection on the first five human chromosomes, across grids of fixed mutation
and selection parameters. We then combined these into a synthetic genome, and
overlaid mutations on the ARG. Then, we ran our likelihood methods on the
resulting allele count data to assess model accuracy. We ran additional
synthetic genome simulations like these to evaluate the impact of two model
violations: recessivity of deleterious mutations and expanding populations. For
the latter, after $9.3$ generations, we grew the population by factor of 1.004
each generation to mimic the human expansion out-of-Africa
\parencite{Gutenkunst2009-pg}. We did not simulate population bottlenecks since
our analyses showed little difference between bottlenecked out-of-Africa
samples (CEU and CHB) and YRI samples. More details about these and the segment
simulations shown in Figure \ref{fig:figure-1}A-C can be found Supplementary
Materials Section \ref{supp:simulation}.
